---
title: "RMangleSeedlings"
author: "Lila Uzzell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(plyr)
library(ggplot2)
library(colorspace)
library(scales)
```
#Using fishers exact test for R. Mangle Developmental Stage
##This file is updated from an R Scrip file called "seedlings.R". The section used from that file is the "06/28/24" on line 46. A Fishers exact test will be run instead of a chi-squared test because the expected frequency is less than 5. Data will need to be manipulated from origial file to fix mistakes and change NULL values to NA. 
```{r data manipulation}
#Change the capital "S" in 'seedling' in row 15, column 6 to lowercase. Change your NULL values to NA.
View(seedlings_wide)
seed <- seedlings_wide
  seed[seed == "NULL"] = NA
  seed[15,6] = "seedling"
View(seed)

#Set the working directory to your "data" folder and export the updated table to a csv.
setwd("~/Documents/UVI/Thesis/Statistics/R_mangle_Seedlings/RMangleSeedlings/data")
write.csv(seed, file = "seed.csv", row.names = FALSE)
```

```{r January vs. April}
#Stage class data for R. mangle monitoring from January to April
table(seed$January,seed$April)
fisher.test(seed$January, seed$April)
```

```{r April vs. August}
table(seed$April,seed$August)
fisher.test(seed$April, seed$August)
```

```{r August vs. February}
table(seed$August,seed$February)
fisher.test(seed$August, seed$February)
```

#Using linear models to look at percent change in seedling heights
```{r Looking at Growth Rate Seed Heights}
#Create a new table that consolidates columns for ease of running stats
seed.consolidate <- R_Mangle_Seedling_Data_FULL %>% 
  select(Bay, MoYr, Monitoring_Period, NumberDays, Quadrat, Habitat, Stage, Tag, Tracking_Period, Surv, Surv_Consolidated, Condition, Water_Depth, PercChgOverallByDays)
head(seed.consolidate)
```

```{r Looking at Growth Rate Seed Heights}
#change fields to become factors and numeric
seed.consolidate$MoYr = as.factor(seed.consolidate$MoYr)
seed.consolidate$PercChgOverallByDays = as.numeric(seed.consolidate$PercChgOverallByDays)
seed.consolidate$Habitat = as.factor(seed.consolidate$Habitat)
seed.consolidate$Bay = as.factor(seed.consolidate$Bay)
seed.consolidate$Monitoring_Period = as.character(seed.consolidate$Monitoring_Period)
seed.consolidate$NumberDays = as.factor(seed.consolidate$NumberDays)
seed.consolidate$Quadrat = as.factor(seed.consolidate$Quadrat)
seed.consolidate$Stage = as.factor(seed.consolidate$Stage)
seed.consolidate$Tag = as.factor(seed.consolidate$Tag)
seed.consolidate$Tracking_Period = as.factor(seed.consolidate$Tracking_Period)
seed.consolidate$Surv = as.factor(seed.consolidate$Surv)
seed.consolidate$Surv_Consolidated = as.factor(seed.consolidate$Surv_Consolidated)
seed.consolidate$Condition = as.factor(seed.consolidate$Condition)
View(seed.consolidate)
```

```{r Looking at Growth Rate Seed Heights}
growthrate <- lm(PercChgOverallByDays ~ MoYr*Habitat, data = seed.consolidate)
summary(growthrate)
```

```{r Create Plot}
plot <- ggplot(seed.consolidate, aes(x=MoYr, y= PercChgOverallByDays, fill= Habitat)) +
  geom_boxplot()+
  theme_classic() +
  scale_fill_brewer(palette = "BuPu") +
  xlab("Monitoring Period")+
  ylab("Percent Change in Height Overall By Day")
plot
```


```{r Play around with Bar Graphs}
#lets remove January and the new plants that occurred each visit (or, the plants with percent change of 0)
organize.data <- seed.consolidate %>% 
  select(Bay, Habitat, MoYr, Monitoring_Period, Tracking_Period, Quadrat, Tag, Stage, Surv_Consolidated, PercChgOverallByDays, everything())
View(organize.data)
```

```{r remove january and NA data or Dead plants}
remove.jan.na <- organize.data %>% 
  filter(Surv_Consolidated != "D",
         MoYr != "23-Jan")
View(remove.jan.na)
```

## linear models for Habitat and Bay with zeros included
```{r Looking at % change in Seed Heights between time period and habitat WITH ZERO DATA INCLUDED}
growthrate.nojan.hab <- lm(PercChgOverallByDays ~ MoYr*Habitat, data = remove.jan.na)
summary(growthrate.nojan.hab)
```

```{r Looking at % change in Seed Heights between time period and bay WITH ZERO DATA INCLUDED}
growthrate.nojan.bay <- lm(PercChgOverallByDays ~ MoYr*Bay, data = remove.jan.na)
summary(growthrate.nojan.bay)
```

## linear models for Habitat and Bay with zeros removed
```{r remove zero data}
# Removing zero data removes the new plants found within each monitoring period (i.e.- new plants would always have a percent change in growth of 0.0 because they appeared between monitoring periods)
remove.zero <- remove.jan.na %>% 
  filter(PercChgOverallByDays != "0")
View(remove.zero)
```

```{r Looking at % change in Seed Heights between time period and habitat ZEROS REMOVED}
growthrate.nojan.nozero.hab <- lm(PercChgOverallByDays ~ MoYr*Habitat, data = remove.zero)
summary(growthrate.nojan.nozero.hab)
```

```{r Looking at % change in Seed Heights between time period and bay ZEROS REMOVED}
growthrate.nojan.nozero.bay <- lm(PercChgOverallByDays ~ MoYr*Bay, data = remove.zero)
summary(growthrate.nojan.nozero.bay)
```

## Plots for Percent Change in Seedling Height Overall By Day
```{r color pallets}
pal1 <- sequential_hcl(2, "Mint")
```

###Zeros Included
```{r Create Plot for Habitat with zeros included}
plot.hab <- ggplot(remove.jan.na, aes(x=MoYr, y= PercChgOverallByDays, fill= Habitat)) +
  geom_boxplot()+
  coord_cartesian(ylim = c(-0.2,0.5))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")+
  theme_classic() +
  scale_fill_brewer(palette = "BuPu")+
  xlab("Monitoring Period")+
  ylab("Percent Change in Height Overall By Day")
plot.hab
```

```{r Create Plot for Bay with zeros included}
plot.bay <- ggplot(remove.jan.na, aes(x=MoYr, y= PercChgOverallByDays, fill= Bay)) +
  geom_boxplot()+
  coord_cartesian(ylim = c(-0.2,0.5))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")+
  theme_classic() +
  scale_fill_brewer(palette = "BuPu") +
  xlab("Monitoring Period")+
  ylab("Percent Change in Height Overall By Day")
plot.bay
```

### Zeros Excluded
```{r Create Plot for Habitat with zeros removed}
plot.hab.zeros.removed <- ggplot(remove.zero, aes(x=MoYr, y= PercChgOverallByDays, fill= Habitat)) +
  geom_boxplot()+
  coord_cartesian(ylim = c(-0.2,0.6))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")+
  theme_classic() +
  scale_fill_brewer(palette = "Orange") +
  xlab("Monitoring Period")+
  ylab("Percent Change in Height Overall By Day")
plot.hab.zeros.removed
```

```{r Create Plot for Bay with zeros removed}
plot.bay.zeros.removed <- ggplot(remove.zero, aes(x=MoYr, y= PercChgOverallByDays, fill= Bay)) +
  geom_boxplot()+
  coord_cartesian(ylim = c(-0.2,0.6))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray")+
  theme_classic() +
  scale_fill_brewer(palette = "BuPu") +
  xlab("Monitoring Period")+
  ylab("Percent Change in Height Overall By Day")
plot.bay.zeros.removed
```

##Using real time data to plot change in height over time
```{r new table for adding real time}
add.time <- remove.zero %>%
  mutate(
    event_date = case_when(
      MoYr == "24-Feb" ~ as.Date("2024-02-13"),  # If category is "24-Feb", assign "2024-02-13"
      MoYr == "23-Apr" ~ as.Date("2023-04-03"),  # If category is "23-Apr", assign "2023-04-03"
      MoYr == "23-Aug" ~ as.Date("2023-08-24"),  # If category is "23-Aug", assign "2023-08-24"
      TRUE ~ NA_Date_                           # Default value for unmatched cases
    )
  )

print(add.time)
```


```{r Graphing change in height over time}
# Create a line plot with a trendline
ggplot(add.time, aes(x = event_date, y = PercChgOverallByDays)) +
  geom_smooth(method = "lm", se = FALSE, color = "olivedrab", linetype = "solid") +
  labs(
    title = expression(paste("Average Change in ", italic("R. mangle"), " Seedling Height Over Time")),
    x = "Date",
    y = "Average Change in Height (per day)"
  )+
  theme_classic() +
  scale_y_continuous(
    breaks = pretty_breaks(n=10)
  ) +
  scale_x_date(date_labels = "%b-%Y", date_breaks = "1 month")
```

```{r Using linear regression for change in height over time}
# Fit linear regression model
seedling.lm <- lm(PercChgOverallByDays ~ event_date, data = add.time)

# Summarize the model
summary(seedling.lm)
```
```{r Extract equation}
slope <- coef(seedling.lm)[2]  # Slope
intercept <- coef(seedling.lm)[1] # Intercept
print(slope) #0.0004417564
print(intercept) #-8.596396
# Create the equation as a string
equation <- paste("y =", round(intercept, 3), "+", round(slope, 5), "x")
print(equation)

r_squared <- summary(seedling.lm)$r.squared
p_value <- summary(seedling.lm)$coefficients[2, 4]

# Create the equation with R² and p-value
equation <- paste(
  "y =", round(intercept, 2), "+", round(slope, 5), "x\n",
  "R² =", round(r_squared, 2), "\n",
  "p =", round(p_value, 3))
print(equation)
```

